HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-10-13 docs/HLL_exp_bad/main.tex
small changes to formal analysis

\section{Experimental Results}\label{sec:exp}
Many different implementations of HLL are available online. Among the repositories collecting more than 350 stars on GitHub are ~\cite{clahll,datasketch}. HLL is also featured in many frameworks, including Redis~\cite{redis}, Google's BigQuery~\cite{bigquery}, Facebook's Airlift~\cite{airlift} and several products of Apache such as Spark~\cite{spahll} and Druid~\cite{druhll}. It is interesting to point out that all these implementations use a constant number of buckets and a fixed hash function. Thus they are potentially vulnerable to our attacks.

@@

First, when attacking a sketch already containing 1000 random items, we can see that the adversary is on average able to insert about half of the items in our starting set in scenario S2, and about 83\% in scenario S4. These results align well with our theoretical analysis.

@@

We can also compare our results from Table~\ref{table:tab2} with results from the attack presented in~\cite[Section 5.1]{hllvuln}. In that paper, the authors target the Redis implementation of HLL and perform the attack against an empty sketch filled with adversarial inputs only. Given a set of 250,000 items, the adversary could choose ``74,390 distinct items and obtain an HLL estimate of only 15,780''~\cite{hllvuln}, achieving a five-fold reduction from the true cardinality after 4 iterations (resets). As explained in Section~\ref{sec:redis}, the Redis implementation analysed in~\cite{hllvuln} does not conform precisely to the original HLL algorithm as presented in Figure~\ref{fig:hll}. Also as discussed in Section~\ref{sec:redis}, the Redis implementation has changed its HLL algorithm several times in recent releases, but the authors of~\cite{hllvuln} do not specify which version was used in their experiments. This makes it difficult to offer a precise comparison with their results. Nevertheless, the S1 adversary following our strategy is able to add 977 and 97,433 distinct items while holding the cardinality estimate at 1 and 126, respectively. Compared to~\cite{hllvuln}, we can reduce the cardinality estimate by a factor of almost 1,000 instead of 5. This comes at the cost of performing more resets compared to~\cite{hllvuln}.

@@

\subsection{Scenario S1}
We begin by disposing of scenario S1. In that scenario, the attacker has access to insertion and reset capabilities, and to cardinality estimates, for a shadow HLL that is identical in every respect to the target HLL. The attacks we presented previously are oblivious to any internals of the implementation (other than that they follow the HLL description in Figure~\ref{fig:hll}). This means that, say, replacing the hash function $h$ with a keyed function or any other cryptographic primitive will be futile in preventing attacks. This illustrates the powerful nature of the resettable shadow device (despite our attacks in that setting being more complex and requiring more effort than in scenarios S2 and S4). In short, we cannot hope to gain any security in scenario S1 by adding cryptography.

@@

@@

We consider adversary ${\cal A}$ to be an algorithm running against a challenger ${\cal C}$ in a security game. The game begins with ${\cal C}$ selecting a bit $b$ uniformly at random. Adversary ${\cal A}$ interacts with up to four oracles which are handled by ${\cal C}$; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\cal A}$ returns a bit $b'$ estimating whether it is interacting with a real HLL instantiation ($b'=0$) or an ideal one ($b'=1$). We define ${\cal A}$'s S$i$ Real-or-Ideal advantage against HLL to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny HLL}}({\cal A}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$. 
The four oracles that  ${\cal A}$ can interact with are as follows:
\begin{itemize}
\item \texttt{init}: this oracle performs initialisation of the HLL (and selection of the PRF key when $h$ is replaced by a PRF $f$); this oracle is available in all of S2, S3, S4, but can only be called once as ${\cal A}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {\cal X}$, the oracle inserts value $x$ into the HLL. This oracle is available in all of S2, S3, S4. Without loss of generality (since the HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made to \texttt{insert($\cdot$)} are distinct. We let $q$ denote the number of insertion queries made by ${\cal A}$ during the course of its attack.
\item \texttt{read}: this oracle returns the HLL cardinality estimate. We allow multiple queries to this oracle and make it available to ${\cal A}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents. This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; we allow multiple queries here because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

We now replace the hash function $h$ acting on ${\cal X}$ with a keyed function $F: \{0,1\}^k \times {\cal X} \rightarrow  \{0,1\}^{n+\ell}$. Here the first input to $F$ is a $k$-bit string, the key, denoted $K$. We assume that \texttt{init} in the HLL security model now initialises key $K$ by selecting it uniformly at random from $\{0,1\}^k$.

@@

The interpretation of the above theorem is that if $F$ is a good PRF, then $\mbox{Adv}^{\mbox{\tiny PRF}}_{F}({\cal B})$ should be small for any adversary ${\cal B}$ running in a reasonable amount of time. Hence we can conclude that the advantage of any S4 adversary against HLL using PRF $F$ is also small.

\begin{proof}
PRF adversary ${\cal B}$ against $F$ is built as follows. Let $d$ denote the hidden bit determining whether ${\cal B}$ receives real responses of the form $F(K,x)$ where $K$ is a uniformly random key ($d=0$) or responses from a truly random function ($d=1$); ${\cal B}$ runs S4 adversary ${\cal A}$ and responds to each of its queries as follows:
\begin{itemize}
\item \texttt{init}: here ${\cal B}$ initialises an HLL filter in accordance with Figure~\ref{fig:hll}. Thus it initialises an array $M[j]$ in which each entry is set to 0. 
\item \texttt{insert($\cdot$)}: on input $x \in {\cal X}$,  ${\cal B}$ makes a query on $x$ to its real-or-random oracle and receives value $y$; ${\cal B}$ uses $y$ to update the HLL bucket $j$ generated via $j=1 + <y_1y_2...y_{n}>_2$ an in the regular HLL algorithm (shown in Figure~\ref{fig:hll}).
\item \texttt{read}: ${\cal B}$ uses its knowledge of the buckets' contents to compute and return the HLL cardinality estimate in accordance with Figure~\ref{fig:hll}. 
\item \texttt{corrupt}: this oracle returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents. 
\end{itemize}
When ${\cal A}$ outputs a bit $b'$, ${\cal B}$ outputs its own bit $d'=b'$. Note that when $d=0$, ${\cal B}$ plays against the real $F$ with a random key $K$ and provides a simulation for ${\cal A}$ that perfectly instantiates the real world. On the other hand, when $d=1$, ${\cal B}$ plays against a truly random function and provides a simulation for ${\cal A}$ that perfectly instantiates the ideal world. The equality of advantages in the statement of the theorem now follows immediately. The remainder of the proof follows on counting queries and noting that the HLL algorithm for computing cardinality is efficient, so ${\cal B}$'s running time when handling \texttt{read} queries is not increased significantly compared to that of ${\cal A}$.
\end{proof}

@@

Finally, we conclude our analysis by reiterating the key point: we have shown that no efficient S4 adversary ${\cal A}$ can tell that it is playing against HLL implemented using truly random values $y$ instead of with values obtained from a PRF, except with a small advantage. This means that ${\cal A}$ is effectively reduced to attacking HLL in the exact same setting used in the average case analysis of HLL from~\cite{hll}, wherein it is proved that the HLL cardinality estimate endures relative errors of the form $1.04/\sqrt{m}$.

>>>>

\section{Experimental Results}\label{sec:exp}
Many different implementations of HLL are available online. Among the repositories collecting more than 350 stars on GitHub are~\cite{clahll,datasketch}. HLL is also featured in many frameworks, including Redis~\cite{redis}, Google's BigQuery~\cite{bigquery}, Facebook's Airlift~\cite{airlift} and several products of Apache such as Spark~\cite{spahll} and Druid~\cite{druhll}. It is interesting to point out that all these implementations use a constant number of buckets and a fixed hash function. Thus they are potentially vulnerable to our attacks.

@@

First, when attacking a sketch already containing 1000 random items, we can see that the adversary is on average able to insert about half of the $2^{22.8}$ items in our starting set in scenario S2, and about 83\% in scenario S4. The cardinality estimate is, on average, increased slightly in S2 (from 1,000 to 1,011) and remains at the same starting value (1,005 on average) in S4. In both cases, we inserted more than 3 million items whilst increasing the cardinality estimate by about $0.1\%$!  These results align well with our theoretical analysis.

@@

We can also compare our results from Table~\ref{table:tab2} with results from the attack presented in~\cite[Section 5.1]{hllvuln}. In that paper, the authors target the Redis implementation of HLL and perform the attack against an empty sketch filled with adversarial inputs only. Given a set of 250,000 items, the adversary could choose ``74,390 distinct items and obtain an HLL estimate of only 15,780''~\cite{hllvuln}, achieving a five-fold reduction from the true cardinality after 4 iterations (resets). As explained in Section~\ref{sec:redis}, the Redis implementation analysed in~\cite{hllvuln} does not conform precisely to the original HLL algorithm as presented in Figure~\ref{fig:hll}. Also as discussed in Section~\ref{sec:redis}, the Redis implementation has changed its HLL algorithm several times in recent releases, but the authors of~\cite{hllvuln} do not specify which version was used in their experiments. This makes it difficult to make a precise comparison with their results. Nevertheless, the S1 adversary following our strategy against the classical HLL algorithm is able to add 977 and 97,433 distinct items while holding the cardinality estimate at 1 and 126, respectively. Compared to~\cite{hllvuln}, we can reduce the cardinality estimate by a factor of almost 1,000 instead of 5. This comes at the cost of performing more resets compared to~\cite{hllvuln}.

@@

\subsection{Scenario S1}
We begin by disposing of scenario S1. In that scenario, the attacker has access to insertion and reset capabilities, and to cardinality estimates, for a shadow HLL that is identical in every respect to the target HLL. The attacks we presented previously are oblivious to any internals of the implementation (other than that they follow the HLL description in Figure~\ref{fig:hll}). This means that, say, replacing the hash function $h$ with a keyed function or any other cryptographic primitive would be futile in preventing attacks in scenario S1. Interestingly, our attacks in scenario S1 are more complex and require more effort than in scenarios S2 and S4. The fact that we cannot prove security in S1 with improved cryptography is not in contradiction to this, but rather an illustration of the power conferred by giving the adversary access to an exact and resettable replica of the target device in the keyed setting.

In summary, we cannot hope to gain any security in scenario S1 by adding more cryptography.

@@

We note that our formal security analysis could be conducted using the framework recently laid out in~\cite{CCS:ClaPatShrPS19}. However, we prefer to give a compact and self-contained treatment here.

@@

We consider adversary ${\cal A}$ to be an algorithm running against a challenger ${\cal C}$ in a security game. The game begins with ${\cal C}$ selecting a bit $b$ uniformly at random. Adversary ${\cal A}$ interacts with up to four oracles which are handled by ${\cal C}$; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\cal A}$ returns a bit $b'$ estimating whether it is interacting with a real HLL instantiation ($b'=0$) or an ideal one ($b'=1$). For $i=1, 2, 3$, we define ${\cal A}$'s S$i$-Real-or-Ideal advantage against HLL to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny HLL}}({\cal A}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$. 
The four oracles that  ${\cal A}$ can interact with are as follows:
\begin{itemize}
\item \texttt{init}: this oracle performs initialisation of the HLL (and selection of the PRF key when $h$ is replaced by a PRF $F$); this oracle is available in all of S2, S3, S4, but can only be called once as ${\cal A}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {\cal X}$, the oracle inserts value $x$ into the HLL. This oracle is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\cal A}$ to \texttt{insert($\cdot$)} are distinct. We let $q$ denote the number of insertion queries made by ${\cal A}$ during the course of its attack.
\item \texttt{read}: this oracle returns the HLL cardinality estimate. We allow multiple queries to this oracle and make it available to ${\cal A}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; we allow multiple queries here because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

We now replace the hash function $h$ acting on ${\cal X}$ from the HLL algorithm with a keyed function $F: \{0,1\}^k \times {\cal X} \rightarrow  \{0,1\}^{n+\ell}$. Here the first input to $F$ is a $k$-bit string, the key, denoted $K$. We assume that \texttt{init} in the HLL security model now initialises key $K$ by selecting it uniformly at random from $\{0,1\}^k$.

@@

The interpretation of the above theorem is that if $F$ is a good PRF, then $\mbox{Adv}^{\mbox{\tiny PRF}}_{F}({\cal B})$ should be small for any adversary ${\cal B}$ running in a reasonable amount of time. Hence we can conclude that the advantage of any S4 adversary  ${\cal A}$ against HLL using PRF $F$ is also small.

\begin{proof}
PRF adversary ${\cal B}$ against $F$ is built as follows. Let $d$ denote the hidden bit determining whether ${\cal B}$ receives real responses of the form $F(K,x)$ where $K$ is a uniformly random key ($d=0$) or responses from a truly random function ($d=1$); ${\cal B}$ runs S4 adversary ${\cal A}$ and responds to each of its queries as follows:
\begin{itemize}
\item \texttt{init}: here ${\cal B}$ initialises an HLL filter in accordance with Figure~\ref{fig:hll}. Thus it initialises an array $M[j]$ in which each entry is set to 0. 
\item \texttt{insert($\cdot$)}: on input $x \in {\cal X}$,  ${\cal B}$ makes a query on $x$ to its real-or-random oracle and receives value $y$; ${\cal B}$ uses $y$ to update the HLL bucket $j$ generated via $j=1 + <y_1y_2...y_{n}>_2$ as in the regular HLL algorithm (shown in Figure~\ref{fig:hll}).
\item \texttt{read}: ${\cal B}$ uses its knowledge of the buckets' contents to compute and return the HLL cardinality estimate in accordance with Figure~\ref{fig:hll}. 
\item \texttt{corrupt}: ${\cal B}$ returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents. 
\end{itemize}
When ${\cal A}$ outputs a bit $b'$, ${\cal B}$ outputs its own bit $d'=b'$. Note that when $d=0$, ${\cal B}$ plays against the real $F$ with a random key $K$ and provides a simulation for ${\cal A}$ that perfectly instantiates the real world. On the other hand, when $d=1$, ${\cal B}$ plays against a truly random function and provides a simulation for ${\cal A}$ that perfectly instantiates the ideal world. The equality of advantages in the statement of the theorem now follows immediately. The remainder of the proof follows on counting queries and noting that the HLL algorithm for computing cardinality is efficient, so ${\cal B}$'s running time due to handling \texttt{read} queries is not increased significantly compared to that of ${\cal A}$.
\end{proof}

@@

Finally, we conclude our analysis by reiterating the key point: we have shown that no efficient S4 adversary ${\cal A}$ can tell that it is playing against HLL implemented using truly random values $y$ instead of with values obtained from a PRF, except with a small advantage. This means that ${\cal A}$ is effectively reduced to attacking HLL in the exact same setting used in the average case analysis of HLL from~\cite{hll}, wherein it is proved that the HLL cardinality estimate endures relative errors of the form $1.04/\sqrt{m}$. It does not matter that ${\cal A}$ can adaptively choose its inputs $x$ in \texttt{insert($\cdot$)} queries, nor that it can access \texttt{read} and \texttt{corrupt} queries. For each \texttt{insert($\cdot$)} query, a fresh value $x$ leads to a new random $y$ being generated and used to update a bucket, as in the average case analysis of HLL.
